Reproducing "Fluctuation Domains

### 2020-02-07 rescience/article.Rmd
Rendered article using ReScience LaTeX templates & fonts :pen:

---
title: 'Reproducing "Fluctuation Domains"'
author: "Carl Boettiger"
date: "1/14/2020"
output: pdf_document
---

Computational reproducibility is

@@

@@

The same stategy worked on MacOSX (Catalina) with `gsl` installed via Homebrew. I was able to update the original autoconf script by recalling that it had been adapted from another package, the `gsl` R package <https://cran.r-project.org/web/packages/gsl/>, which also binds against the GSL library. Working from the autoconf files provided there, I could successfully get the package to configure and compile.  This highlights a potential risk in changes that are intended to make the package more portable but also add complexity.  More generally, while the package has no dependency on external R packages (beyond those default packages provided as part of base R: `stats`, `utils`, `graphics`, and `grDevices`, this issue illustrates the familiar reproducibility challenge that is created by changing versions of dependencies and system libraries (GSL in this case).  System libraries against which a package is compiled are particularly pernicious, because versioning of these components is both platform dependent and separate from the R package manager process, and thus cannot be addressed merely with package version tracking in tools such as `packrat` [@packrat] or `renv`.  Today, containerized approaches such a Docker [see @Boettiger2015; @Boettiger2017] provide a solution to this system-dependency issue that was not available a decade ago.

The second challenge was in figuring out just what commands had to be run to duplicate the figures.  As mentioned above, the GitHub version of the source-code did not have the manuscript source, and even the version that did have the mansucript source did not write the manuscript itself as a dynamic document (e.g. in Sweave [@Sweave]).  In that version, files `fig1.R` and `fig2.R` are were included in the same subdirectory as the `pdf` figures referenced by the `tex` version of the manuscript, but `fig2.R`, but these files contained only the code required to actually generate the figures themselves once the result data had been pre-computed and saved as `.txt` files. That result data files had also not been archived seperately.  Fortunately, re-reading my own documentation, I was able to deduce that these text files were generated by running the function `logistic()` with the parameter settings given in the caption of Fig 2 in the manuscript.

@@

```{r}
library(fluctuationDomains)
## Figure 1 function definitions:
logistic_curve = function(x){-x*exp(-x*x/2)}
chemostat_curve <- function(x, d = 0.1, q = 0.1){
  (q*x-d/(1-x*d))* (d/x-2*d**2)
}
branching_curve <- function(x, SIGMA2_K = 2, SIGMA2_C = 1){
  -exp( (x^2/2)*(4/SIGMA2_C -1/SIGMA2_K) ) *
    (SIGMA2_C+exp(2*x^2/SIGMA2_C)*SIGMA2_C-2*SIGMA2_K)*x /
    ( (1+exp(2*x^2/SIGMA2_C)^2)*SIGMA2_C*SIGMA2_K)
}

@@

Figure 2 has the core computational results demonstrating the consequences of these fluctuation regimes.  The code for this analysis was implemented as pure C code, which can be run in stand-alone mode and generates three output text files.  The R package uses the `.C` interface to pass parameters to the C function (`Rlogistic`) which runs the ensemble and writes its output as text. The C function does not expose a random seed, so re-running the analysis creates results which are not bitwise identical each time. Of course the scientific results are expected to be independent of the seed anyway, so it is perhaps reassuring that as long as the ensemble of replicates is large enough, the same pattern can be reproduced quite closely. The original manuscript uses ensembles of 100,000 replicates, which I have reproduced here.  A possible source of confusion is that the code defaults to only 100 replicates if the ensemble size is not explicitly verified -- which is an ensemble small enough to still see random deviations from the predicted results.   I probably set that lower default to facilitate faster testing.  Likewise, the MAXTIME parameter has to be extended from the default setting to allow the third initial condition set to run sufficiently long.

```{r}
out = logistic(Xo = 1,  ENSEMBLES = 10^5, MAX_TIME = 5000)
```

```{r}
out = logistic(Xo = 2,  ENSEMBLES = 10^5, MAX_TIME = 5000)
```

```{r}
out = logistic(Xo = 3,  ENSEMBLES = 10^5, MAX_TIME = 10000)
```

Note that in the modern standard for R packages with methods in implemented in C, several things would be different from my original implementation.  The C methods are expected to bind R's random number generator engines, and not use their own methods, which allows them to coordinate with standard R methods for setting seed and so forth.  Also, the data should be passed by reference back to R and not serialized to disk (though this can introduce memory constraints).  Third, it is now also standard practice to register the C routines explicitly.  The built-in CHECK methods of R will flag these issues.

While the package is organized as an R package, it is worth noting that it is also not written up to the standards required of modern R packages to pass the built-in `R CMD check` testing routines successfully.  Some of these are not particularly material

- C routines are not registered
- `SystemRequirements` field was not provided in the package `DESCRIPTION`.  By adding this to list `libgsl`, it is possible to have the package successfully install on the `rhub` [@rhub] testing platforms.
- The package does not use the R random number generator.  This was intentional, as the package was written as stand-alone C code that happened to have an R interface, though notably even the C implementation fails to expose a random number `seed`. 
- missmatch in documentation and function definition

```{r}
```

>>>>

---
title: 'Reproducing "Fluctuation Domains"'
output: 
  pdf_document:
    latex_engine: xelatex
    template: rescience.latex
    includes:
      before_body: header.tex
      after_body: suffix.tex
    keep_tex: true
documentclass: rescience
bibliography: "bibliography.bib"

---

@@

In this analyis I seek to reproduce the numerical results of [@Boettiger2010], which focuses on the evolutionary trajectory of species trait under selection, as predicted by the so-called "Canonical Equation" of adaptive dynamics [@Dieckmann1996].  Theoretical work I present in the paper predicts the emergence of a 'flutation enhancement' regime far from an evolutionary stable point, and suggests that the evolution of traits which start in this distant region would deviate substantially from the expected evolutionary path predicted by the canonical equation.  In particular, the distribution of such trajectories could be bimodal, making the "canonical" path far from the most likely path.  The details of this distribution cannot be solved for analytically however, so in this paper I resorted to an replicate exact stochastic simulations of the evolutionary model using a Gillespie algorithm [@Gillespie].

I implemented Gillespie algorithm for the evolutionary model in C code, and wrapped the C code in an R package structure to make it easier to run and easier to plot results.  The C code depended on one standard C library, the GNU Scientific Library, which has evolved considerably (with one major version change) since that time.  The original source code was 'published' (initially only on a colleague's website, as stated in the paper, though later copies were submitted to formal archives, as discussed below).  That original copy was distributed under a GPL v3 licence, though I later replaced that with a CC0 license when submitting the version to a data archive (Dryad) which required public domain declarations.

@@

The same stategy worked on MacOSX (Catalina) with `gsl` installed via Homebrew. I was able to update the original autoconf script by recalling that it had been adapted from another package, the `gsl` R package <https://cran.r-project.org/web/packages/gsl/>, which also binds against the GSL library. Working from the autoconf files provided there, I could successfully get the package to configure and compile.  This highlights a potential risk in changes that are intended to make the package more portable but also add complexity.  More generally, while the package has no dependency on external R packages (beyond those default packages provided as part of base R: `stats`, `utils`, `graphics`, and `grDevices`, this issue illustrates the familiar reproducibility challenge that is created by changing versions of dependencies and system libraries (GSL in this case).  System libraries against which a package is compiled are particularly pernicious, because versioning of these components is both platform dependent and separate from the R package manager process, and thus cannot be addressed merely with package version tracking in tools such as `packrat` [@packrat] or `renv` [@renv].  Today, containerized approaches such a Docker [see @Boettiger2015; @Boettiger2017] provide a solution to this system-dependency issue that was not available a decade ago.

The second challenge was in figuring out just what commands had to be run to duplicate the figures.  As mentioned above, the GitHub version of the source-code did not have the manuscript source, and even the version that did have the mansucript source did not write the manuscript itself as a dynamic document (e.g. in Sweave [@knitr_book]).  In that version, files `fig1.R` and `fig2.R` are were included in the same subdirectory as the `pdf` figures referenced by the `tex` version of the manuscript, but `fig2.R`, but these files contained only the code required to actually generate the figures themselves once the result data had been pre-computed and saved as `.txt` files. That result data files had also not been archived seperately.  Fortunately, re-reading my own documentation, I was able to deduce that these text files were generated by running the function `logistic()` with the parameter settings given in the caption of Fig 2 in the manuscript.

@@

```{r fig1, fig.cap="The three panels of figure 1, showing examples of evolutionary landscapes and the corresponding fluctation domains for three classic models (logistic, chemostat, branching model). This figure is largely theoretical, and so easily replicates Figure 1 of the original paper.  The first panel showing the logistic model depicts the fluctation regimes for the model parameters used in the numerical analysis."}
library(fluctuationDomains)
## Figure 1 function definitions:
logistic_curve = function(x){-x*exp(-x*x/2)}
chemostat_curve <- function(x, d = 0.1, q = 0.1){
  (q*x-d/(1-x*d))* (d/x-2*d**2)
}
branching_curve <- function(x, SIGMA2_K = 2, SIGMA2_C = 1){
  -exp( (x^2/2)*(4/SIGMA2_C -1/SIGMA2_K) ) *
    (SIGMA2_C+exp(2*x^2/SIGMA2_C)*SIGMA2_C-2*SIGMA2_K)*x /
    ( (1+exp(2*x^2/SIGMA2_C)^2)*SIGMA2_C*SIGMA2_K)
}

@@

Figure 2 of the original paper has the core computational results demonstrating the consequences of these fluctuation regimes.  The code for this analysis was implemented as pure C code, which can be run in stand-alone mode and generates three output text files.  The R package uses the `.C` interface to pass parameters to the C function (`Rlogistic`) which runs the ensemble and writes its output as text. The C function does not expose a random seed, so re-running the analysis creates results which are not bitwise identical each time. Of course the scientific results are expected to be independent of the seed anyway, so it is perhaps reassuring that as long as the ensemble of replicates is large enough, the same pattern can be reproduced quite closely. The original manuscript uses ensembles of 100,000 replicates, which I have reproduced here.  A possible source of confusion is that the code defaults to only 100 replicates if the ensemble size is not explicitly verified -- which is an ensemble small enough to still see random deviations from the predicted results.   I probably set that lower default to facilitate faster testing.  Likewise, the MAXTIME parameter has to be extended from the default setting to allow the third initial condition set to run sufficiently long.

The code was never parallelized, but runs with a small memory footprint by streaming simulation data out (e.g. to disk) over `fprint`.  Replicating all results takes about an hour of computational time on a Intel-i7 processor.

```{r fig.cap = "A starting condition of Xo = 1, at the edge of the fluctuation dissipation regime.  The whole evolutionary trajectory thus occurs within the dissipation regime, and mean dynamics closely match that of the canonical equation, and show a tight normal distribution around the expected path.  This matches the trajectories seen figure 2 of the original manuscript."}
out = logistic(Xo = 1,  ENSEMBLES = 10^5, MAX_TIME = 3000)
```

```{r fig.cap="A starting condition of Xo = 2, matching the trajectories seen figure 2 of the original manuscript. Deviations from the canonical path are visible, but the distribution is not yet bimodal."}
out = logistic(Xo = 2,  ENSEMBLES = 10^5, MAX_TIME = 5000)
```

```{r fig.cap="Starting deep in the predicted fluctuation enhancement regime, Xo = 3, the resulting distribution of trajectories is bimodal, just as shown in the original figure 2 of the original manuscript."}
out = logistic(Xo = 3,  ENSEMBLES = 10^5, MAX_TIME = 10000)
```

### On the R package compendium

My original paper acknowledges the use of the R package compendium approach [@TempleLang2007] for reproducible research, and idea later developed in @Marwick.  While a decade ago the package as created would pass R's built-in `R CMD check` routine, standards have since evolved and the current implementation would not do so:

- The C methods are expected to bind R's random number generator engines, and not use their own methods, which allows them to coordinate with standard R methods for setting seed and so forth.   
- Data should be passed by reference back to R and not serialized to disk (though this can introduce memory constraints).  
- It is now also standard practice to register the C routines explicitly.  
- `SystemRequirements` field was not provided in the package `DESCRIPTION`.  By adding this to list `libgsl`, it is possible to have the package successfully install on the `rhub` [@rhub] testing platforms.

Note that these tighter integrations would also prevent the C libraries from running as purely stand-alone implementation.  In any event, meeting all the automated checks of a formal package is not an expectation of the compendium approach, which merely borrows some of the package tooling.

A dynamic document using RMarkdown [@rmarkdown] for reproducing the present mansucript can be found in the `rescience` sub-directory which I have now added to the GitHub repository housing the original R package, <https://github.com/cboettig/fluctuationDomains>, git tag `rescience-submission`, now archived with DOI <https://doi.org/xxxx>.

# Conclusions

In conclusion, I have been able to successfully replicate the original results.  As the original results were not performed with a fixed random seed, replicated results are statistically comparable but not bitwise identical.  In a scientific sense this is perhaps more useful than replication with a fixed seed.  My compendium approach did not install entirely out of the box, as an autoconf configure script first required updating.  Ironically, the original version of the code published on a co-author's personal website and linked from the manuscript both still existed and did install out-of-the-box.  Reproducing the results of the paper required some careful reading of parameters listed in the paper and manually knowing what functions to call, as this paper did not include a complete "dynamic document" [@knitr; @knitr_book; @rmarkdown] approach discussed in @Marwick and also used in this replication paper.  I think this experience underscored the critical role dependency management has in frustrating reproducibility: this replication was relatively simple in part due to the low number of dependencies and the technical simplicity of the code.  The other take-away for me is how easy it is to leave out small but important steps needed to reproduce a figure when not using the dynamic document approach.  Despite having R scripts called "figure1.R" and "figure2.R", these scripts could not simply be run by themselves to create the figures.

# References
